A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-08-01 paper-submitted-2.tex
Updates from Overleaf

\begin{keypoints}
\item Power system coefficients derived from a time series of GIC and geoelectric field measurements are frequency dependent.
\item A GIC model with frequency-dependent system coefficients provides significantly better estimates of GIC than a frequency independent model.
\item Data-derived power system coefficients may differ significantly with those computed using power system configuration information.
\end{keypoints}

\begin{abstract}
  A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are computed from data and are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system components. It is also shown that further improvements over Model~2 can be made using frequency-dependent models with the geomagnetic field as an input.
\end{abstract}

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system coefficients $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system coefficients are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model. Two additional frequency-dependent models are considered that require only the geomagnetic field as an input and they are shown to be significantly better at estimating GIC than the models that use only the geoelectric field as an input.

@@

GIC was measured in a grounded neutral point of a Y-connected three-phase transformer connected to the 187 kV bus at the Memanbetsu substation of the Hokkaido Power Co. The 187 kV line extends from the Ashro to Memanbetsu substation, which are both grounded. Memanbetsu is an end point of the line.

@@

\noindent
the matrix equation $\underline{\mathbf{GIC}} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is solved in a least squares sense, where $\underline{\mathbf{GIC}}$ is a zero mean 86400x1 matrix with rows of $GIC(t)$; $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ with zero mean columns. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent closed-form equations for solving the matrix equation.]{Pulkkinen2007} The notation $\overline{a}_o$ and $\overline{b}_o$ will be used for the average of the $a_o$ and $b_o$ values computed using the 35 1-day window segments of available data.

@@

\begin{linenomath*}
  \begin{equation}
    G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
    \label{model2}
  \end{equation}
\end{linenomath*}
\noindent
and were estimated using the Fourier transforms of  $GIC(t)$ and $\mathbf{E}(t)$; the method of coefficient estimation and the generation of time-domain GIC estimates for this and the following models are given at the end of this section.

Motivated by the results described in the following section, in which Model~2 provides a significantly better representation of GIC than Model~1, we have also considered two additional models that can be used to estimate GIC. In these two models, a direct measurement of the geoelectric field is not used; instead, the driver is the geomagnetic field, which is much more frequently measured, and as a result, these models may be of practical interest.

In Model~3, the coefficients $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field, $\mathbf{B}$, and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$ and

@@

\noindent
where, explicitly, the electric field components are $E'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $E'_y = Z_{yx}B_x + Z_{yy}B_y$.
That is, instead of using the measured electric field directly as in Model~2, a magnetotelluric transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$ and $\mathbf{B}(t)$ and no electric field measurements are used; this model is expected to produce estimates of GIC that are similar to that obtained from Model~3 because they both use the same input measurements.

@@

The same approach is used for solving for $z_x$ and $z_y$ in Equation~\ref{model4}. For the transfer function $\boldsymbol{Z}$ in $\mathbf{E}(\omega) = \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$ used in Model~3, the $xx$ and $xy$ components of $\boldsymbol{Z}$ are solved for using

@@

To interpolate the frequency-domain transfer function components estimated at the evaluation frequencies, their real and imaginary components were linearly interpolated individually onto the original DFT frequency grid (with a spacing of $1/86400$~Hz), and then the inverse DFT was computed. The interpolated transfer function coefficients were set to zero at frequencies higher than $0.375$~Hz and the transfer function estimate at $f=0$ was set to zero before interpolation.

GIC time series were calculated by multiplication of the interpolated transfer function coefficients with the DFT of the 24-hr long and 1~s cadence model inputs and computing the inverse DFT (e.g, for Model~2, $G_E(f)=a^I(f)E_x(f)+b^I(f)E_y(f)$ was calculated, where the superscript $I$ indicates the interpolated value, and then the inverse DFT of $G_E(f)$ was calculated to give the estimate of $G_E(t)$ on a 1~s time grid).

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for gaussian-distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods longer than $\sim$30~s; At periods shorter than $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to a low signal-to-noise ratio.)

@@

The mean-squared-error for model $m$ on day $D$ is
\begin{equation*}
  \mbox{MSE}_m = \frac{1}{N}\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2 \,,
\end{equation*}
\noindent
where $GIC_m(t)$ is the model prediction of model $m$ at time step $t$, $t_o=600$, and $t_f=85800$, so that sum is over one day of one-second-cadence measurements excluding the first and last 10~minutes. The MSE ratio for model $m$ is MSE$_1$/MSE$_m$.

The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is calculated for day $D$ using
\begin{equation*}
  \mbox{PE}_m = 1 - \frac{\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2}{\sum_{t=t_o}^{t_f} \big(GIC(t)-\overline{GIC(t)}\big)^2}\,,
\end{equation*}
\noindent
where $\overline{GIC(t)}$ is an average over the same time interval used in the summations. The PE is a measure of the skill of the model relative to the variance in the data that are predicted and it represents the fraction of the variance in the data that is predicted by the model.

The average PE and the MSE ratios were computed by averaging the out-of-sample PEs and MSE ratios. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using the bootstrap method \citep{Zoubir1998}: 1000 samples of the out-of-sample prediction metrics were used to compute the CIs. (The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.)

@@

The PE and MSE ratios are shown in Table~\ref{results}. Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~\ref{results}, i.e., Model~1 has the lowest PE and Model~4 has the highest. Note that the large errors visible at the start and end of the time intervals are due to the transients discussed in Section~\ref{section:Model_Evaluation} and the first and last 10~minutes of each interval were omitted from metrics calculations.

\begin{table}
  \caption{Out-of-sample model performance metrics for Model $m$. MSE$_1$/MSE$_m$ is the MSE ratio of Model~1 with respect to that of Model $m$. The PE and MSE ratios are the average of 35 values and the confidence interval, CI, was determined from 1000 bootstrap samples.}
  \centering
  \begin{tabular}{l c c c c}
    \hline
    $m$\hspace{1em} Model & PE & 95\% CI & MSE$_1$/MSE$_m$ & 95\% CI\\
    \hline
    1\hspace{1em} $G_o(t) = a_oE_x(t) + b_oE_y(t)$ & 0.35 & [0.25, 0.45] & 1.0 & \\
    2\hspace{1em} $G_E(t) = a(\omega)E_x(t) + b(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6, 2.0]\\
    3\hspace{1em} $G_{E'}(t) = a(\omega)E'_x(t) + b(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8, 3.4]\\
    4\hspace{1em} $G_{B}(t) = z_xB_x(t) + z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8, 5.3]\\
    \hline
  \end{tabular}
  \label{results}
\end{table}

The primary conclusions of this work follow from the results shown in Table~\ref{results}: the frequency dependent model, Model~2, is significantly better at estimating GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

@@

Figure~\ref{SN} shows the average signal-to-noise (i.e., signal to prediction error) ratios as a function of period for the four models. At each evaluation frequency, the SN ratio is the ratio of the average of the out-of-sample SN ratios and the error bars represent a 95\% CIs based on 1000 bootstrap samples of the segment SN ratios. Consistent with the results in Table~\ref{results}, the SN ratio for Model~1 is lower than that for Models~2-4 at all frequencies. However, ordering of the SN ratio for Models~2-4 is dependent on period.

@@

Above $10,000$~s (2.8~hrs), the SN ratio ordering is the same as the ordering of the PE and MSE ratios shown in Table~\ref{results}, with Model~1 having the lowest and Model~4 having the largest SN ratio.

@@

Figure~\ref{Z} shows that the frequency domain transfer functions for Model~2 vary by an order of magnitude on timescales of 60~s to 12~h (that for Model~1 is constant by definition). For this system, GIC is more sensitive to $E_y$ than $E_x$ due to $b(\omega) > a(\omega)$, but note that $E_y(t)$ has less variability than $E_x(t)$ - the variance in $E_x(t)$ is $\sim$5x larger than $E_y(t)$ over the entire dataset (this is also visible in Figure~\ref{sample}). Note that transfer function shown in Figure~\ref{Z} differs significantly from that of a layered slab conductivity model for which $|Z_{xy}|=|Z_{yx}|$.

The ratio of $|\overline{b}_o|/|\overline{a}_o|$ is approximately 1 for Model~1 and this can be compared with $|b(\omega)|/|a(\omega)|$ calculated using the lines for $|a(\omega)|$ and $|b(\omega)|$ shown in Figure~\ref{Z}. The ratio varies from 2.7 at $T=30$~s to 1.6 at $T=12$~h. 
Figure~\ref{Phi} shows the transfer function coefficient phase angles (that for Model~1 is 0$^{\circ}$ for positive coefficients and 180$^{\circ}$ for negative coefficients by definition). The phases for $a(\omega)$ and $b(\omega)$ are frequency dependent. At $T=12$~h, the phases of $a(\omega)$ and $b(\omega)$ differ by 180$^{\circ}$ and the associated signs are consistent with those of $\overline{a}_o > 0$ and $\overline{b}_o < 0$. At $T=12$~h, the values of $b(\omega)$ and $a(\omega)$ and Model~2 correspond to an electric field of 60$^{\circ}$ west of north being associated with the largest GIC.

Both Model~1 and 2 predict a maximum GIC when the electric field angle is significantly different from the angle of the 187~kV line. This can be caused by part other parts of the power network (the details of which are unavailable) making significant contributions to GIC and variations in ground conductivity.

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memanbetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer function magnitudes differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function parameters were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code used was LEMI MT V1.4), suggests the explanation is not due to an error in the code used here to compute transfer functions. We expect to consider this in a future investigation as it suggests that stack averaging can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

In the discussion of (3) above, we have attempted to address independently the contribution of spatial variations in the electric field magnitude and direction and spatial variations in the electric field spectra, but generally these variations do not occur independently as they are both caused by horizontal variations in the ground conductivity. Thus, the finding that the angle of the electric field that produces maximum GIC differs significantly from the angle of the 187~kV power line for both Model~1 and Model~2 is likely to be consistent with the frequency dependence of $a$ and $b$ found in Model~2. However, testing this claim would require a full network analysis and knowledge of the ground conductivity throughout the network.

@@

\acknowledgments
Magnetic and electric field measurements made at the Kakioka Magnetic Observatory were obtained from the Japan Meteorological Agency Data Portal \url{http://www.kakioka-jma.go.jp/obsdata/metadata/en}. The geomagnetic coordinates in Figure~\ref{map} were computed using code from \url{https://www.ngdc.noaa.gov/geomag/geom_util/apex.shtml}. We would like to thank the Hokkaido Electric Power Co. for the GIC measurements at the Memanbetsu substation and Shinichi Watari for communicating and preparing the GIC data and providing feedback on the manuscript. We also thank Ikuko Fujii for providing clarifications on the geoelectric instrument response, Trevor Gaunt and Emanuel Benabeau for review and comments on part (1) of Section~\ref{discussion}, and Josh Rigler for substantial feedback and review.
%\clearpage

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/map.pdf}
  \caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memambetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
  \label{map}
\end{figure}

>>>>

\begin{keypoints}
\item Power system coefficients derived from a time series of GIC and geoelectric field measurements are frequency dependent.
-\item Data-derived power system coefficients may differ significantly with those computed using power system configuration information.
\end{keypoints}

\begin{abstract}
  A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system components. It is also shown that further improvements over Model~2 can be made using frequency-dependent models with the geomagnetic field as an input.
\end{abstract}

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system coefficients $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system coefficients are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model. Two additional frequency-dependent models are considered that require only the geomagnetic field as an input and they are shown to be significantly better at estimating GIC than the models that use the geoelectric field as an input.

@@

GIC was measured in a grounded neutral point of a Y-connected three-phase transformer connected to the 187 kV bus at the Memanbetsu substation of the Hokkaido Power Co. The 187 kV line extends from the Ashro substation to the Memanbetsu substation, which are both grounded. Memanbetsu is an end point of the 187 kV line.

@@

\noindent
the matrix equation $\underline{\mathbf{GIC}} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is solved in a least squares sense, where $\underline{\mathbf{GIC}}$ is a zero mean 86400x1 matrix with rows of $GIC(t)$; $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ with zero mean columns. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent analytic equations for solving the matrix equation.]{Pulkkinen2007} The notation $\overline{a}_o$ and $\overline{b}_o$ will be used for the average of the $a_o$ and $b_o$ values computed using 35 1-day-window-segment regressions.

@@

\begin{linenomath*}
  \begin{equation}
    G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
    \label{model2}
  \end{equation}
\end{linenomath*}
\noindent
and were estimated using the Fourier transforms of  $GIC(t)$, $E_x(t)$, and $E_y(t)$; the method of coefficient estimation and the generation of time-domain GIC estimates for this and the following models are given at the end of this section.

Motivated by the results described in the following section, in which Model~2 provides a significantly better representation of GIC than Model~1, we have also considered two additional models that can be used to estimate GIC. In these two models, a direct measurement of the geoelectric field is not used; instead, the driver is the geomagnetic field, which is much more often measured, and as a result, these models may be of practical interest.

In Model~3, previously computed coefficients $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field, $\mathbf{B}$, and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$ and

@@

\noindent
where, explicitly, the electric field components are $E'_x(\omega)$ = $Z_{xx}(\omega)B_x(\omega) + Z_{xy}(\omega)B_y(\omega)$ and $E'_y(\omega) = Z_{yx}(\omega)B_x(\omega) + Z_{yy}(\omega)B_y(\omega)$.
That is, instead of using the measured electric field directly as in Model~2, a magnetotelluric transfer function that relates $\mathbf{E}(\omega)$ to $\mathbf{B}(\omega)$ is used to provide an estimate of the electric field based on the magnetic field.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$, $B_x(t)$, and $B_y(t)$ and no electric field measurements are used; this model is expected to produce estimates of GIC that are similar to that obtained from Model~3 because they both use the same input measurements.

@@

The same approach is used for solving for $z_x$ and $z_y$ in Equation~\ref{model4}.

For the transfer function $\boldsymbol{Z}$ in $\mathbf{E}(\omega) = \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$ used in Model~3, the $xx$ and $xy$ components of $\boldsymbol{Z}$ are solved for using

@@

To interpolate the frequency-domain transfer function components that are estimated at the evaluation frequencies onto the original DFT frequency grid (with a spacing of $1/86400$~Hz), their real and imaginary components were linearly interpolated individually. The interpolated transfer function coefficients were set to zero at frequencies higher than $0.375$~Hz and the transfer function estimate at $f=0$ was set to zero before interpolation.

GIC time series were calculated by multiplication of the interpolated transfer function coefficients with the DFT of the 24-hr long and 1~s cadence model inputs and computing the inverse DFT (e.g, for Model~2, $G_E(\omega)=a^I(\omega)E_x(\omega)+b^I(\omega)E_y(\mega)$ was calculated, where the superscript $I$ indicates the interpolated value, and then the inverse DFT of $G_E(\omega)$ was calculated to give the estimate of $G_E(t)$ on a 1~s time grid).

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with either a Huber weighting using a constant of $1.345$ or bi-square weighting using a constant of $4.685$ (both with no hard-cutoff step). The commonly used constants provide 95\% efficiency for gaussian-distributed errors. Models were also computed using time domain pre-whitening (using a 10th-order Yule-Walker model computed using the model output), time domain tapering (using a Parzen window), and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional procedures individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal-to-noise ratio is above $\sim$3 (corresponding to periods longer than $\sim$30~s); at periods shorter than $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to a low signal-to-noise ratio.)

@@

The mean-squared-error for model $m$ on day $D$ is
\begin{equation*}
  \mbox{MSE}_m = \frac{1}{N}\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2 \,,
\end{equation*}
\noindent
where $GIC_m(t)$ is the estimate from model $m$ at time step $t$;  $t_o=600$ and $t_f=85800$, so that sum is over one day of one-second-cadence measurements excluding the first and last 10~minutes. The MSE ratio for model $m$ is MSE$_1$/MSE$_m$.

The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is calculated for day $D$ using
\begin{equation*}
  \mbox{PE}_m = 1 - \frac{\sum_{t=t_o}^{t_f} \big(GIC(t)-GIC_m(t)\big)^2}{\sum_{t=t_o}^{t_f} \big(GIC(t)-\overline{GIC(t)}\big)^2}\,,
\end{equation*}
\noindent
where $\overline{GIC(t)}$ is an average over the same time interval used in the summations. The PE is a measure of the skill of the model with respect to the variance in the predictand and it represents the fraction of the variance in the data that is predicted by the model (when PE~$\ge 0)$.

The average PE and the MSE ratios were computed by averaging the out-of-sample PEs and MSE ratios. The 95\% confidence interval, CI, on the PE and MSE ratios were calculated using the bootstrap method \citep{Zoubir1998}: 1000 random samples (with replacement) of 17 of the 35 out-of-sample prediction metrics were used to compute the CIs. The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.

@@

The PE and MSE ratios are shown in Table~\ref{resultstable}. Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~\ref{resultstable}, i.e., Model~1 has the lowest PE and Model~4 has the highest. Note that the large errors visible at the start and end of the time intervals are due to the transients discussed in Section~\ref{section:Model_Evaluation} and the first and last 10~minutes of each interval were omitted from metrics calculations.

\begin{table}
  \caption{Out-of-sample model performance metrics for Model $m$. MSE$_1$/MSE$_m$ is the MSE ratio of Model~1 with respect to that of Model $m$. The PE and MSE ratios are the average of 35 values and the confidence interval, CI, was determined from 1000 bootstrap samples.}
  \centering
  \begin{tabular}{l c c c c}
    \hline
    $m$\hspace{1em} Model & PE & 95\% CI & MSE$_1$/MSE$_m$ & 95\% CI\\
    \hline
    1\hspace{1em} $G_o(t) = a_oE_x(t) + b_oE_y(t)$ & 0.35 & [0.25, 0.45] & 1.0 & \\
    2\hspace{1em} $G_E(t) = a(\omega)E_x(t) + b(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6, 2.0]\\
    3\hspace{1em} $G_{E'}(t) = a(\omega)E'_x(t) + b(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8, 3.4]\\
    4\hspace{1em} $G_{B}(t) = z_xB_x(t) + z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8, 5.3]\\
    \hline
  \end{tabular}
  \label{resultstable}
\end{table}

The primary conclusions of this work follow from the results shown in Table~\ref{resultstable}: the frequency dependent model, Model~2, is significantly better at estimating GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

@@

Figure~\ref{SN} shows the average signal-to-noise (i.e., signal to prediction error) ratios as a function of period for the four models. At each evaluation frequency, the SN ratio is the ratio of the average of the out-of-sample SN ratios and the error bars represent a 95\% CIs based on 1000 bootstrap samples of the segment SN ratios. Consistent with the results in Table~\ref{resultstable}, the SN ratio for Model~1 is lower than that for Models~2-4 at all frequencies. However, ordering of the SN ratio for Models~2-4 is dependent on period.

@@

Above $10,000$~s (2.8~hrs), the SN ratio ordering is the same as the ordering of the PE and MSE ratios shown in Table~\ref{resultstable}, with Model~1 having the lowest and Model~4 having the largest SN ratio.

@@

Figure~\ref{Z} shows that the frequency domain transfer functions for Model~2 vary by an order of magnitude on timescales of 60~s to 12~h (that for Model~1 is constant by definition). For this system, GIC is more sensitive to $E_y$ than $E_x$ due to $b(\omega) > a(\omega)$, but note that $E_y(t)$ has less variability than $E_x(t)$ - the variance in $E_x(t)$ is $\sim$5x larger than $E_y(t)$ over the entire dataset (this is also visible in Figure~\ref{sample}).

The ratio of $|\overline{b}_o|/|\overline{a}_o|$ is approximately 1 for Model~1 and this can be compared with $|b(\omega)|/|a(\omega)|$ calculated using the lines for $|a(\omega)|$ and $|b(\omega)|$ shown in Figure~\ref{Z}. The ratio varies from 2.7 at $T=30$~s to 1.6 at $T=12$~h.

Figure~\ref{Phi} shows the transfer function coefficient phase angles (that for Model~1 is 0$^{\circ}$ for positive coefficients and 180$^{\circ}$ for negative coefficients by definition). The phases for $a(\omega)$ and $b(\omega)$ are frequency dependent. At $T=12$~h, the phases of $a(\omega)$ and $b(\omega)$ differ by 180$^{\circ}$ and the associated signs are consistent with those of $\overline{a}_o > 0$ and $\overline{b}_o < 0$. At $T=12$~h, the values of $b(\omega)$ and $a(\omega)$ and Model~2 correspond to an electric field of 60$^{\circ}$ west of north being associated with the largest GIC.

Both Model~1 and~2 predict a maximum GIC when the electric field angle is significantly different from the angle of the 187~kV line. As discussed in the following section, this can be caused by other parts of the power network (the details of which are unavailable) making significant contributions to GIC and variations in ground conductivity. 
Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that transfer function shown in Figure~\ref{Z_MT} differs significantly from that of a layered slab conductivity model for which $|Z_{xy}|=|Z_{yx}|$. That electric and magnetic field data from Memanbetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015} and the fact that we did not use a remote reference in our calculations. After 2006 (January through May tested) our calculated transfer function magnitudes differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function parameters were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and results from extensive verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code used was LEMI MT V1.4), suggests the explanation is not due to an error in the code used to compute the transfer functions in this work. We expect to consider this in a future investigation as it suggests that stack averaging can be a significant source of transfer function bias and expect that the explanation involves non-stationarity in the noise.

@@

In the discussion of (3) above, we have attempted to address independently the contribution of spatial variations in the electric field magnitude and direction and spatial variations in the electric field spectra, but generally these variations do not occur independently as they are both caused by horizontal variations in the ground conductivity. Thus, the finding that the angle of the electric field that produces maximum GIC differs significantly from the angle of the 187~kV power line for both Model~1 and Model~2 is likely to be consistent with the frequency dependence of $a$ and $b$ found in Model~2. However, testing this claim would require a full network analysis and knowledge of the ground conductivity throughout the network (\cite{Uyeshima2001} has shown significant variations in the impedance tensor $\mathb{Z}$ on Hokkaido Island).

@@

\acknowledgments
Magnetic and electric field measurements made at the Kakioka Magnetic Observatory were obtained from the Japan Meteorological Agency Data Portal \url{http://www.kakioka-jma.go.jp/obsdata/metadata/en}. The geomagnetic coordinates in Figure~\ref{map} were computed using code from \url{https://www.ngdc.noaa.gov/geomag/geom_util/apex.shtml}. The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]. We thank the Hokkaido Electric Power Co. for the GIC measurements at the Memanbetsu substation and Shinichi Watari for communicating and preparing the GIC data and providing feedback on the manuscript. We also thank Ikuko Fujii for providing clarifications on the geoelectric instrument response, Trevor Gaunt and Emanuel Benabeau for review and comments on part (1) of Section~\ref{discussion}, and Josh Rigler for substantial feedback and review.
%\clearpage

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/map.pdf}
  \caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memanbetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
  \label{map}
\end{figure}
